# Answers to practical exercises

Below are all the questions and answers to the practical exercises at the end of each exercise. 

## Diversity indices
In the folder there is another (invented) dataset using British bats called `bat-communities.csv`. Read in the data, manipulate it as required by `vegan`, then answer the following questions.

```{r, message = FALSE}
# Load packages
library(vegan)
library(picante)

# Read in the dataset
bat <- read.csv("data/bat-communities.csv")
# Create a matrix we can use with vegan
bat.matrix <- sample2matrix(bat)
```

1. Which site has the fewest species?
```{r}
specnumber(bat.matrix)
```

*Site B*

2. How many different species are there in total?
```{r}
# How many unique species are there?
length(unique(bat$Species[bat$Abundance >0]))
```

*18*

3. What is Simpson's diversity index for Site J?
```{r}
# Simpson's index
diversity(bat.matrix, index = "simpson")
```

*0.8198242*

4. Draw a species accumulation curve for the bats and estimate the total number of species. If you round up numbers with decimal places, what is the maximum number of species estimated by any metric?
```{r}
# Fit species accumulation curve
bat.curve <- specaccum(bat.matrix, method = "random", permutations = 1000)

# Plot the curve
plot(bat.curve, ci.type = "poly", col = "blue", ci.col = "lightblue", 
     lwd = 2, ci.lty = 0, xlab = "number of sites", 
     ylab = "cumulative number of bat species")

# Estimate diversity
specpool(bat.matrix)
```

*18.267 + 0.476, i.e. 19 species in total*

## Phylogenies
Read in the `frog-tree.nex` phylogeny from the folder. This comes from @feng2017phylogenomics. Then do the following:

Read the tree into R.
```{r, message = FALSE}
# Load packages
library(ape)
library(ggtree)
library(tidyverse)

# Read in the tree
tree <- read.nexus("data/frog-tree.nex")

# Look at it
tree
```

1. Use R functions to determine: 
a) how many species are in the tree?
```{r}
# Look at the tree structure
str(tree)
# Or the whole tree
tree
```    

*There are 214 tips*

b) is the tree fully resolved?
```{r}
# Check whether the tree is binary/fully resolved
is.binary.tree(tree)
```

*Yes*

c) is the tree rooted?
```{r}
# Check whether the tree is rooted
is.rooted(tree)
```    

*Yes*

2. Use `plot.phylo` to plot the tree.
    a) Can you change the size of the tip labels?
    b) Can you make a fan-shaped plot? 
    c) Can you change the colour of the tips and branches?
```{r}
# Plot the tree
plot(tree, 
     no.margin = TRUE, # to make sure it fills the whole space
     type = "fan", # to make a fan tree
     cex = 0.6, # to reduce the tip label size (default = 1)
     tip.color = "deeppink", # to change the tip colour
     edge.color = "springgreen") # to change the branch colour
# yuk!
```

3. Save the tree to file as "mysuperdoopertree.nex"    
```{r, eval = FALSE}
# Save the tree
write.nexus(tree, file = "data/mysuperdoopertree.nex")
``` 

## Preparation
In the data folder there is another tree (`primate-tree.nex`) and dataset (`primate-data.csv`) for investigating the evolution of primate life-history variables. These data come from the PanTHERIA database [@jones2009pantheria] and 10kTrees [@arnold201010ktrees]. 

Read in the tree and data then prepare them for a PCM analysis.
```{r}
# Load packages
library(ape)
library(geiger)
library(tidyverse)
library(treeplyr)

# Read in the tree
primatetree <- read.nexus("data/primate-tree.nex")
# Look at the tree summary
str(primatetree)

# Plot the tree as a circular/fan phylogeny with small labels
plot(primatetree, cex = 0.2, typ = "fan", no.margin = TRUE)

# Check whether the tree is binary
# We want this to be TRUE
is.binary.tree(primatetree) 

# It is FALSE so make the tree binary:
primatetree <- multi2di(primatetree)

# Check whether the tree is binary now...
is.binary.tree(primatetree) 

# Check whether the tree is rooted
# We want this to be TRUE
is.rooted(primatetree)

# Check whether the tree is ultrametric
# We want this to be TRUE, but it is FALSE
is.ultrametric(primatetree)
# Force ultrametric
primatetree <- force.ultrametric(primatetree)

# Read in the data
primatedata <- read_csv("data/primate-data.csv")

# Look at the data
glimpse(primatedata)

# Replace spaces with underscores in species names
primatedata <- 
  primatedata %>%
  mutate(Binomial = str_replace(string = Binomial, pattern = " ", replacement = "_"))
# Check it worked
glimpse(primatedata)

# Check whether the names match in the data and the tree
check <- name.check(phy = primatetree, data = primatedata, 
                    data.names = primatedata$Binomial)
# Look at check
check

# 149 species are in the tree but not the data
# 0 species are in the data but not the tree
  
# Combine and match the tree and data
primatestuff <- make.treedata(tree = primatetree,  data = primatedata, 
                              name_column = "Binomial")

# Look at the tree summary
primatestuff$phy
# Look at the data
glimpse(primatestuff$dat)
## There are 77 species in the tree and the data

# Make a new column called tiplabel with the tip labels in it
primatestuff$dat$tiplabel <- primatestuff$phy$tip.label

# Save tree as mytree
mytree <- primatestuff$phy
# Save data as mydata
mydata <- primatestuff$dat

# Force mydata to be a data frame
mydata <- as.data.frame(mydata)
# Check mydata is now a dataframe
str(mydata)

## OPTIONAL
# Make any required subsets of the tree/data

## OPTIONAL
# Write the cleaned data to a new file
# write_csv(mydata, path = "data/clean-primate-data.csv")

# Write the cleaned tree to a new file
# write.nexus(mytree, file = "data/clean-primate-tree.nex")
```

## Phylogenetic signal
In the data folder there is another tree (`primate-tree.nex`) and dataset (`primate-data.csv`) for investigating the evolution of primate life-history variables. These data come from the PanTHERIA database [@jones2009pantheria] and 10kTrees [@arnold201010ktrees]. 

Read in the tree and data, then prepare them for a PCM analysis (you may have already done this in the previous exercise which should save you some time). Then use `phylosig` to answer the following questions:

```{r, message = FALSE}
# Load the packages
library(ape)
library(geiger)
library(treeplyr)
library(caper)
library(tidyverse)
library(phytools)

### Prepare the tree and the data as shown in the Preparation answer above ###
### This should mean you now have mydata and mytree ready to work with ###
```

1. What is $\lambda$ for log gestation length?
```{r}
# First create logGestation containing just log gestation length values
logGestation <- log(pull(mydata, GestationLen_d))
# Give log Eye names = species names at the tips of the phylogeny
names(logGestation) <- mydata$tiplabel
# Look at the first few rows
head(logGestation)

# Estimate lambda
lambdaGestation <- phylosig(mytree, logGestation, method = "lambda", test = TRUE)
# Look at the results
lambdaGestation
```

*0.948*

2. What is *K* for log gestation length?
```{r}
# Estimate Blomberg’s *K*
KGestation <- phylosig(mytree, logGestation, method = "K", test = TRUE, nsim = 1000)
# Look at the results
KGestation
```

*0.776*

3. What is D for social status?
```{r}
# Prepare the data for use in caper
primate <- comparative.data(phy = mytree, data = mydata, 
                            names.col = tiplabel, vcv = TRUE, 
                            na.omit = FALSE, warn.dropped = TRUE)

# Force the phylogeny to have polytomies again nrather than zero length branches
primate$phy <- di2multi(primate$phy)

# Estimate D
Dsocial <- phylo.d(data = primate, names.col = tiplabel, binvar = SocialStatus, 
                  permut = 1000)

# Look at the output
Dsocial
```

*~ -0.72, this is a randomisation test so may vary*

## PGLS
In the data folder there is another tree (`primate-tree.nex`) and dataset (`primate-data.csv`) for investigating the evolution of primate life-history variables. These data come from the PanTHERIA database [@jones2009pantheria] and 10kTrees [@arnold201010ktrees]. 

Let's investigate the relationship between gestation length in Primates and their body size. Theory leads us to believe that bigger mammals have longer gestation lengths (compare mice to elephants), but is this true in Primates?

Read in the tree and data, then prepare them for a PCM analysis (you may have already done this in a previous exercise which should save you some time). Fit a PGLS model to investigate the relationship between log gestation length (`y = log(GestationLen_d)`) and log body size (`x = log(AdultBodyMass_g)`) in Primates. Don't forget to look at the model diagnostics! 
```{r, message = FALSE}
# Load the packages
library(ape)
library(geiger)
library(treeplyr)
library(caper)
library(tidyverse)

### Prepare the tree and the data as shown in the Preparation answer above ###
### This should mean you now have mydata and mytree ready to work with ###

# Prepare the data for use in caper
primate <- comparative.data(phy = mytree, data = mydata, 
                            names.col = tiplabel, vcv = TRUE, 
                            na.omit = FALSE, warn.dropped = TRUE)

# Check the list of dropped species is what you expected
primate$dropped$tips
primate$dropped$unmatched.rows

# Fit a PGLS model
model.pgls <- pgls(log(GestationLen_d) ~ log(AdultBodyMass_g), 
                   data = primate, lambda = "ML", bounds = list(lambda = c(0.01, 1)))

### Model diagnostics
# Make a plotting window with 4 panes as there are four plots
par(mfrow = c(2, 2))
# Now plot the model diagnostics
plot(model.pgls)
# Return the plot window to one pane for later plotting
par(mfrow = c(1, 1))

# Look at the model outputs
summary(model.pgls)

# Plot the results
ggplot(mydata, aes(x = log(AdultBodyMass_g), 
                     y = log(GestationLen_d))) +
  geom_point() +
  geom_abline(slope = coefficients(model.pgls)[2], 
              intercept = coefficients(model.pgls)[1]) +
  theme_bw()
```

Then answer the following questions.

1. What is the maximum likelihood estimate of $\lambda$ in the model? 
```{r}
# Look at the model outputs. Lambda is near the middle
summary(model.pgls)
```

*0.892*

2. What are the confidence intervals on the maximum likelihood estimate of $\lambda$. How would you interpret these?
```{r}
# Extract the confidence intervals on lambda
pgls.confint(model.pgls, "lambda")$ci.val
```

*0.753434, 0.966543. We are 95% certain that lambda is between these two values.*

3. Plot the $\lambda$ profile for the the maximum likelihood estimate of $\lambda$. What does it look like?
```{r}
# Create a likelihood profile of the lambda estimate
lambda.profile <- pgls.profile(model.pgls, "lambda")
# Plot the likelihood profile
plot(lambda.profile)
```

4. Is there a significant relationship between log gestation length and log body size? What is the slope of this relationship? 
```{r}
# Look at the model outputs. 
summary(model.pgls)
```

*Yes. p < 0.001. Slope = 0.105 $\pm$ 0.019*

## Models of evolution with continuous traits
In the data folder there is another tree (`primate-tree.nex`) and dataset (`primate-data.csv`) for investigating the evolution of primate life-history variables. These data come from the PanTHERIA database [@jones2009pantheria] and 10kTrees [@arnold201010ktrees]. 

Read in the tree and data, then prepare them for a PCM analysis (you may have already done this in a previous exercise which should save you some time). Fit Brownian and OU models to log adult body mass for Primates (`AdultBodyMass_g`). Then answer the following questions.

```{r}
# Load the packages
library(ape)
library(geiger)
library(treeplyr)
library(caper)
library(tidyverse)

### Prepare the tree and the data as shown in the Preparation answer above ###
### This should mean you now have mydata and mytree ready to work with ###

# Create logBodyMass containing just log body mass values
logBodyMass <- log(pull(mydata, AdultBodyMass_g))
# Give log body mass names = species names at the tips of the phylogeny
names(logBodyMass) <- mydata$tiplabel
# Look at the first few rows
head(logBodyMass)
# Fit the Brownian model
BM <- fitContinuous(mytree, logBodyMass, model = c("BM"))
# Fit the OU model
OU <- fitContinuous(mytree, logBodyMass, model = c("OU"))
```

1. Which model fits best? 
```{r}
# Could look at AIC weights...
aic.scores <- setNames(c(BM$opt$aic, OU$opt$aic), c("BM","OU"))
aicw(aic.scores)

# Or just the raw AIC values
BM$opt$aic
OU$opt$aic
```

*BM but only by a small amount*

2. What is the rate of evolution for the Brownian model?
```{r}
# Look at the whole output
BM
# Or just the one parameter
BM$opt$sigsq
```

*0.028655*

3. What is the strength of pull towards the long-term mean in the OU model? What does this value suggest?
```{r}
# Look at the whole output
OU
# Or just the one parameter
OU$opt$alpha
```

*8.080092e-49, i.e. very close to zero which makes it very close to a Brownian model*

## Models of evolution with discrete traits
In the data folder there is another tree (`primate-tree.nex`) and dataset (`primate-data.csv`) for investigating the evolution of primate life-history variables. These data come from the PanTHERIA database [@jones2009pantheria] and 10kTrees [@arnold201010ktrees]. 

Read in the tree and data, then prepare them for a PCM analysis (you may have already done this in a previous exercise which should save you some time). Fit equal rates, symmetric rates and all rates different models to the social status variable for Primates (`SocialStatus`). Note that social status has two states: non-social (1) or social (2).

```{r}
# Load the packages
library(ape)
library(geiger)
library(treeplyr)
library(caper)
library(tidyverse)

### Prepare the tree and the data as shown in the Preparation answer above ###
### This should mean you now have mydata and mytree ready to work with ###

# Create Social containing just social status values
Social <- pull(mydata, SocialStatus)
# Give social status names = species names at the tips of the phylogeny
names(Social) <- mydata$tiplabel
# Look at the first few rows
head(Social)
# Fit the ER model
equal <- fitDiscrete(mytree, Social, model = "ER")
# Fit the ARD model
ard <- fitDiscrete(mytree, Social, model = "ARD")
```

Then answer the following questions.

1. Which model fits best? 
```{r}
# Using AIC weights
aic.scores <- setNames(c(equal$opt$aic, ard$opt$aic), 
                       c("equal", "different"))
aicw(aic.scores)

# Or just the raw AIC values
equal$opt$aic
ard$opt$aic
```

*All rates different*

2. What is the transition rate from non-social (1) to social (2), and vice versa, in the all rates different model? What does this mean biologically?

```{r}
# Could look at the model output
ard
# Or plot the Q matrix as network
plot.gfit(ard)
```

*1 to 2 is ~0.009, 2 to 1 is very close to zero. Transitions from social (2) to non-social (1) behaviour are extremely rare.*

3. Plot ancestral state estimates of social status (from the ARD model) on the phylogeny. Is the root node estimated as being non-social or social?
```{r, message = FALSE}
# Ancestral state estimation (ignore the warnings)
ancestral_social <- ace(Social, mytree, type = "discrete", model = "ARD")

# Choose colours
social_colours <- c("deeppink", "goldenrod")

# Plot the tree with colours at the tips and nodes to represent socail status
plot(mytree, cex = 0.5, adj = c(0.2), type = "fan", 
     no.margin = TRUE, show.tip.label = TRUE)
tiplabels(pch = 16, col = social_colours[mydata$SocialStatus])
nodelabels(pie = ancestral_social$lik.anc, piecol = social_colours, cex = 0.5)
legend("bottomleft", pch = 15, bty = "n",
       legend = c("Non Social", "Social"), 
       col = social_colours)
```

*Root is non-social*

4. Why didn't I ask you to fit the symmetric model?

*Because with only two states the symmetric model is equivalent to the equal rates model*

## Beyond simple macroevolutionary models of trait evolution
In the data folder there is another tree (`primate-tree.nex`) and dataset (`primate-data.csv`) for investigating the evolution of primate life-history variables. These data come from the PanTHERIA database [@jones2009pantheria] and 10kTrees [@arnold201010ktrees]. 

Read in the tree and data, then prepare them for a PCM analysis (you may have already done this in a previous exercise which should save you some time). 

Using log body size (`AdultBodyMass_g`) as your continuous trait, and social status (`SocialStatus`) to define the selective regimes, fit (A) a multi-rate Brownian motion model and (B) an OU model with trait optima or long-term mean values for different evolutionary regimes but the same $\sigma^2$ and $\alpha$. Note that social status has two states: non-social (1) or social (2). Assume it is evolving according to an all rates different model.

```{r, message = FALSE}
# Load the packages
library(ape)
library(geiger)
library(treeplyr)
library(caper)
library(tidyverse)
library(OUwie)

### Prepare the tree and the data as shown in the Preparation answer above ###
### This should mean you now have mydata and mytree ready to work with ###

# Create Social containing just social status values
Social <- pull(mydata, SocialStatus)
# Give social status names = species names at the tips of the phylogeny
names(Social) <- mydata$tiplabel

# Simulate node and tip values for 500 trees using the ARD model
trees.ARD <- make.simmap(mytree, Social, model = "ARD", nsim = 500)
# Store the results of the summary of the 500 trees
summary.ARD <- summary(trees.ARD)

# Get the max likelihood state at each node
best <- apply(summary.ARD$ace, 1, which.max)

# Assign to the tree node labels
mytree$node.label <- levels(as.factor(mydata$SocialStatus))[best]

# Make new dataset with just three columns for OUwie
bodymass <- data.frame(species = mydata$tiplabel, 
                    regime = mydata$SocialStatus, 
                    trait = log(mydata$AdultBodyMass_g))

# Fit BMS model
BMS <- OUwie(mytree, bodymass, model = "BMS", root.station = FALSE)

# Fit OUM model
OUM <- OUwie(mytree, bodymass, model = "OUM")
```

Then answer the following questions:

1. Which is the best fitting model?
```{r}
# Extract AICc scores from the models
aic.scores <- setNames(c(BMS$AICc, OUM$AICc), 
                           c("BMS", "OUM"))
# Get aic weights
aicw(aic.scores)
```
*OUM*

2. If you were running these analyses for a publication, what additional models would you compare these models to?

*BM and single peak OU for sure. Possibly one or more of the more complex models if you had a good biological reason to*

3. In the multi rate Brownian model, which social status state has the higher rate of log body mass evolution?
```{r}
# Look at the model output
BMS
```

*Non-social (1) has the higher rate of body mass evolution 0.04031313*

4. Look at the parameter estimates of the multi peak OU model. Do you see anything worrying there? It might help to look at a histogram showing the body sizes of species in the non-social and the social categories.
```{r}
# Look at the output
OUM

# Plot the body sizes for the different groups
ggplot(mydata, aes(x = log(AdultBodyMass_g), fill = as.factor(SocialStatus))) +
  geom_histogram(bins = 20, alpha = 0.5) +
  scale_fill_manual(values = social_colours) +
  facet_wrap(~SocialStatus, ncol = 1) +
  theme(legend.position = "none") +
  theme_bw()
```

*The optima estimates are way off. Non-social species have a long-term mean log body size estimate of 5.918899, which makes sense looking at the histograms of the data. But social species have a long-term mean log body size estimate of 18607801, which if you exponentiate it is infinity. Even King Kong was not infinitely large! Something has clearly gone wrong here, but if we had not looked at the model parameters we would have said this was the best model...It's likely we just don't have enough data to fit a model that is as complex as this, so the parameter optimisations are not reliable.*

## Diversification 
`BAMMtools` has a several inbuilt example datasets. For this practical exercise we will use the cetacean (`whales`; from @steeman2009radiation) and the ray-finned fishes (`fishes`; from @rabosky2013rates) datasets.

Load the whales and fishes trees and BAMM outputs data into R (__there is no `mcmc.fishes` file___).
```{r}
# Load the packages
library(ape)
library(BAMMtools)
library(coda)
library(phytools)
library(viridis)

# Load whales data from BAMMtools
data(whales)
data(mcmc.whales)
data(events.whales)

# Load fishes data from BAMMtools
data(fishes)
data(mcmc.fishes)
data(events.fishes)

# Read in the tree, mcmc, and edata for whales
whales_tree <- whales
whales_mcmc <- mcmc.whales
whales_edata <- getEventData(whales, eventdata = events.whales, burnin = 0.1)

# Read in the tree, mcmc, and edata for fishes
fishes_tree <- fishes
fishes_edata <- getEventData(fishes, eventdata = events.fishes, burnin = 0.1)
```

Then answer the following questions: 

1. How can we tell that the BAMM model for whales has converged? What is the effective sample size for number of shifts?
```{r}
# Take a look at the plot
plot(mcmc$logLik ~ mcmc$generation)

# What is 10% of rows?
burnstart <- floor(0.1 * nrow(mcmc))
# Remove the first 10% from the runs 
postburn <- mcmc[burnstart:nrow(mcmc), ]

# Extract ESS for the number of rate shifts parameter
effectiveSize(postburn$N_shifts)

# Extract ESS for the log likelihood
effectiveSize(postburn$logLik)
```

*The plot looks like a furry caterpillar (!), and effective sample sizes are well over 200. ESS = 1225.609*

2. What is the most common number of rate shifts in the posterior for the whales data?
```{r}
# Look at the posterior for number of shifts
summary(whales_edata)
```

*1*

3. Plot the 95% credible set of rate shift configurations for whales.
```{r}
# Extract 95% credible set of rate shift configurations
whales_css <- credibleShiftSet(whales_edata, expectedNumberOfShifts = 1, 
                        threshold = 5, set.limit = 0.95)
# Look at the summary
summary(whales_css)

# Plot 95 % credible set of rate shift configurations
plot.credibleshiftset(whales_css)
```

4. Plot the mean phylorate plot for fishes.
```{r}
# Plot mean phylorate plot with legend and slighting larger line width (lwd)
plot.bammdata(fishes_edata, legend = TRUE, lwd = 2)
```

5. What are the mean speciation rate and the 90% posterior density for mean speciation rate in fishes?
```{r}
# Extract speciation and extinction rates from the whole posterior
allrates <- getCladeRates(fishes_edata)
# Calculate overall mean speciation rate
mean(allrates$lambda)
# Calculate 90% highest posterior density for speciation rate
quantile(allrates$lambda, c(0.05, 0.95))
```

*Mean = 0.09470252, 5% = 0.09020489, 95% =  0.09960825*